Pretty darn good control: when are approximate solutions better than approximate models

### 2023-01-30 manuscript/manuscript.Rmd
mini changes intro

abstract: |
  Existing methods for optimal control methods struggle to deal with the
  complexity commonly encountered in real-world systems, including
  dimensionality, process error, model bias and heterogeneity.  Instead
  of tackling these complexities directly, researchers have typically
  sought to find exact optimal solutions to simplified models of the
  processes in question. When is the optimal solution to a very
  approximate, stylized model better than an approximate solution to a
  more accurate model? This question has largely gone unanswered owing
  to the difficulty of finding even approximate solutions in the case of
  complex models.  Our approach draws on recent algorithmic and
  computational advances in deep reinforcement learning. These methods
  have hitherto focused on problems in games or robotic mechanics, which
  operate under precisely known rules. We demonstrate the ability for
  novel algorithms using deep neural networks to successfully
  approximate such solutions (the ‘‘policy function’’ or control rule)
  without knowing or ever attempting to infer a model for the process
  itself. This powerful new technique lets us finally begin to answer
  the question. We showcase this approach by applying it to the control
  of a highly non-linear three-variable model for a fishery. Here, the
  reinforcement learning agent discovers an effective simplification of
  the problem to obtain a simple, interpretable control rule. Moreover,
  the policy obtained is both more profitable and more sustainable than
  any constant mortality policy---the standard family of policies considered
  in fishery management.

@@

Much effort has been spent grappling with the complexity of our natural world in contrast to the relative simplicity of the models we use to understand it.
Heroic amounts of data and computation are being brought to bear on developing better, more realistic models of our environments and ecosystems, in hopes of improving our capacity to address the many planetary crises.
 But despite these efforts and advances, we remain faced with the difficult task of figuring out how best to respond to these crises.
While simplified models have historically allowed for exploration of large decision spaces,, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
For instance, Global Circulation Models (GCMs) such as HadOM3 [CITE] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of six emissions scenarios from the IPCC [@riahi2017shared].

@@

In this paper, we focus on a well-developed application of model-based management of the natural world that has long illustrated the trade-offs between model complexity and policy complexity: the management of marine fisheries.
Fisheries management is both an important issue to itself and a rich & frequent test-bed of ecological management more generally.
Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
Fisheries management has  roots in both the fields of _ecosystem management_ and _natural resource economics_.
Both fields might trace their origins to the notion of maximum sustainable yield (MSY), introduced independently by a fisheries ecologist [@Schaefer1954] and the economist [@Gordon1954] in the same year.
From this shared origin, each field would depart from the simplifying assumptions of the Gordon-Schaefer model in divergent ways, leading to different techniques for deriving policies from models.The heart of the management problem is easily understood: a manager seeks to set quotas on fishing that will ensure the long-term profitability and sustainability of the industry.
Mathematical approaches developed over the past century may be roughly divided between these two fields: (A), ecologists, focused on ever more realistic models of the biological processes of growth and recruitment of fish while considering relatively stylized suite of potential management strategies, and (B) economists, focused on far more stylized models of the ecology while exploring a far less constrained set of possible policies.
The economist's approach can be characterized by the mathematics of a Markov decision process [MDP @Clark1973; @Clark1990; @Marescot2013], in which the decision-maker must observe the stock each year and recommend a possible action.
In this approach, the policy space that must be searched is exponentially large -- for a management horizon of T decisions and a space of N actions, the number of possible policies is N^T.
In contrast, fisheries ecologists and ecosystem management typically search a space of policies that does not scale with the time horizon.
Under methods such as "Management Strategy Evaluation" (MSE, [@Punt2016]) a manager identifies a candidate set of "strategies" a priori, and then compares the performance of each strategy over a suite of simulations to determine which strategy gives the best outcome (i.e. best expected utility).
This approach is far more amenable to complex simulations of fisheries dynamics and more closely corresponds to how most marine fishing quotas are managed today in the United States [see stock assessments documented in @RAMLegacyDB].
<!-- Do we say here that these strategies retain the 'constant mortality' intuition of @Schaefer? -->

Recent advances in machine learning may allow us to once again bridge these approaches, while also bringing new challenges of their own.
Novel data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RAMLegacyDB].
Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
However, advances _Deep Reinforcement Learning_, (DRL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari1; @atari2] to fusion reactions [@fusion1; @fusion2] to the remarkable dialog abilities of ChatGPT [@chatgpt].
RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, immense computational costs, and frequent challenges with transfer-ability or reproducibility.
A review of all these issues is beyond our scope but can be found elsewhere [@rl_intro; @millie-1; @millie-2].
Here, though, we will focus on the issue of opacity and interpretability raised by these methods.
In contrast with optimization algorithms currently used in either ecosystem management or resource economics, RL algorithms have no guarantees of or metrics for convergence to an optimal solution.
In general, one can only assess the performance of these black box methods relative to alternatives.

<!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
In this article, we compare the performance of a DRL solution to that of a management strategy evaluation across potential 'constant mortality' policies.\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘code’’ subdirectory. The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.} 
We consider the problem of devising harvest strategies in a three species ecosystem (Figure 1B).
We show that, somewhat surprisingly, the DRL policy closely resembles a *bang-bang* solution (see Text Box 1).
Bang-bang solutions are known to be optimal for a variety of single-species models, however, there is no a-priori reason why they should be optimal for higher-dimensional systems.
The DRL policy drives an oscillation process in the system and, in this way, maintains the long-term sustainability of fishing.
In contrast, constant mortality policies overfish at low populations and underfish (relative to the DRL solution) at high populations.
This leads to a tradeoff when leveraging constant mortality strategies: higher constant mortality rates lead to high profitability at the expense of a higher probability of overfishing to extinction, while lower constant mortality rates attain stability at the expense of profit.

@@

In this section we introduce some models describing the population dynamics of fisheries.
In general, the class of models that appear in this context are *first order finite difference equations* (potentially stochastic difference equations).
For $n$ species, these models have the general form
\begin{align}
  \label{eq:general model}
  \begin{split}
    \Delta X_t &:= X_{t+1} - X_t = f_X(N_t) - M_X X_t\\
    \Delta Y_t &:= X_{t+1} - Y_t = f_Y(N_t) - M_Y Y_t\\
    &\dots
  \end{split}
\end{align}
where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations and $f_i:\mathbb{R}^{n}\to\mathbb{R}$ are arbitrary functions.
Here, $M_i=M_i(N_t)\in\mathbb{R}_+$ is a state-dependent fish mortality arising from harvesting (sometimes referred to as *fishing effort*).
Thus, the term $M_X X_t$ is the total $X$ harvest at time $t$.
We will focus on systems where a single species, $X$ is harvested, i.e. $M_Y = M_Z = \dots = 0$.

@@

Reinforcement learning (RL) is a way of approaching *sequential decision problems* through machine learning.
All applications of RL can be conceptually separated into two parts: an *agent* and an *environment* which the agent interacts with.
That is, the agent takes actions within the environment.
As a result of an agent's action, the environment will transition to a new state and return a numerical *reward* to the agent.\footnote{See Fig. TBD in [@rl-intro-marcus] for a conceptual description of reinforcement learning algorithms.}
The rewards encode the agent's goal.
The main task of any RL algorithm is then to maximize the cumulative reward received. 
This objective is achieved by aggregating experience in what is called the *training* period and learning from such experience.
<!-- ML: We could cite our deep RL paper here instead of going into further details. Not sure what we are expecting as the audience. -->.

@@

```{r}
#| label: tuning
#| fig.cap : "Result of the tuning procedure for the constant mortality strategy. 
#| (A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point. Vertical lines correspond to the optimal constant mortality rate $M^* = $ 0.045 (green) and $0.8 M^*=$ 0.036 (violet). 
#| (B) Several constant mortality policies shown for illustrative purposes. Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $M^*X$ shown in green, while $0.8M^*X$ shown in violet. (TBD choose nicer and plentier  slopes)" 
#| out.width : "6.5in"
msy_rewards <- msy_df |>
  group_by(action, rep) |>
  filter(t == max(t)) |> 
  group_by(action) |>
  summarise(mean_reward = mean(reward),
            sd = sd(reward))

@@

In Fig. \ref{fig:state-space}, we perform a back to back visualization of all the policies evaluated up to now---MSY constant mortality, 80% MSY constant mortality, DRL and escapement.
The policies are computed at each state along the paths they traverse in the 50 evaluation rounds, and the values are projected on the $X$-$Z$ plane for visualization.
This figure highlights the similarity between the policy found by the DRL agent and the optimal escapement policy.

>>>>

abstract: |
  Existing methods for optimal control struggle to deal with the
  complexity commonly encountered in real-world systems, including
  dimensionality, process error, model bias and heterogeneity.  Instead
  of tackling these complexities directly, researchers have typically
  sought to find exact optimal solutions to simplified models of the
  processes in question. When is the optimal solution to a very
  approximate, stylized model better than an approximate solution to a
  more accurate model? This question has largely gone unanswered owing
  to the difficulty of finding even approximate solutions in the case of
  complex models.  Our approach draws on recent algorithmic and
  computational advances in deep reinforcement learning. These methods
  have hitherto focused on problems in games or robotic mechanics, which
  operate under precisely known rules. We demonstrate the ability for
  novel algorithms using deep neural networks to successfully
  approximate such solutions (the ‘‘policy function’’ or control rule)
  without knowing or ever attempting to infer a model for the process
  itself. This powerful new technique lets us finally begin to answer
  the question. We showcase this approach by applying it to the control
  of a highly non-linear three-variable model for a fishery. Here, the
  reinforcement learning agent discovers an effective simplification of
  the problem to obtain a simple, interpretable control rule. Moreover,
  the policy obtained is both more profitable and more sustainable than
  any constant mortality policy---the standard family of policies considered
  in fishery management.

@@

Much effort has been spent grappling with the complexity of our natural world in contrast to the relative simplicity of the models we use to understand it.
Heroic amounts of data and computation are being brought to bear on developing better, more realistic models of our environments and ecosystems, in hopes of improving our capacity to address the many planetary crises.
 But despite these efforts and advances, we remain faced with the difficult task of figuring out how best to respond to these crises.
While simplified models have historically allowed for exploration of large decision spaces, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
For instance, Global Circulation Models (GCMs) such as HadOM3 [CITE] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of six emissions scenarios from the IPCC [@riahi2017shared].

@@

In this paper, we focus on a well-developed application of model-based management of the natural world that has long illustrated the trade-offs between model complexity and policy complexity: the management of marine fisheries.
Fisheries management is both an important issue to itself and a rich & frequent test-bed of ecological management more generally.
Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
Fisheries management has  roots in both the fields of _ecosystem management_ and _natural resource economics_.
Both fields might trace their origins to the notion of maximum sustainable yield (MSY), introduced independently by a fisheries ecologist [@Schaefer1954] and the economist [@Gordon1954] in the same year.
From this shared origin, each field would depart from the simplifying assumptions of the Gordon-Schaefer model in divergent ways, leading to different techniques for deriving policies from models.The heart of the management problem is easily understood: a manager seeks to set quotas on fishing that will ensure the long-term profitability and sustainability of the industry.
Mathematical approaches developed over the past century may be roughly divided between these two fields: (A) ecologists, focused on ever more realistic models of the biological processes of growth and recruitment of fish while considering relatively stylized suite of potential management strategies, and (B) economists, focused on far more stylized models of the ecology while exploring a far less constrained set of possible policies.
The economist's approach can be characterized by the mathematics of a Markov decision process [MDP @Clark1973; @Clark1990; @Marescot2013], in which the decision-maker must observe the stock each year and recommend a possible action.
In this approach, the policy space that must be searched is exponentially large -- for a management horizon of T decisions and a space of N actions, the number of possible policies is $N^T$.
In contrast, fisheries ecologists and ecosystem management typically search a space of policies that does not scale with the time horizon.
Under methods such as "Management Strategy Evaluation" (MSE, [@Punt2016]) a manager identifies a candidate set of "strategies" a priori, and then compares the performance of each strategy over a suite of simulations to determine which strategy gives the best outcome (i.e. best expected utility).
This approach is far more amenable to complex simulations of fisheries dynamics and more closely corresponds to how most marine fishing quotas are managed today in the United States [see stock assessments documented in @RAMLegacyDB].
<!-- Do we say here that these strategies retain the 'constant mortality' intuition of @Schaefer? -->

Recent advances in machine learning may allow us to once again bridge these approaches, while also bringing new challenges of their own.
Novel data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RAMLegacyDB].
Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
However, advances _Deep Reinforcement Learning_, (DRL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari1; @atari2] to fusion reactions [@fusion1; @fusion2] to the remarkable dialog abilities of ChatGPT [@chatgpt].
RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, requiring immense computational costs, and presenting frequent challenges with transfer-ability or reproducibility.
A review of all these issues is beyond our scope but can be found elsewhere [@rl-intro; @millie-1; @millie-2].
Here, though, we will focus on the issue of opacity and interpretability raised by these methods.
In contrast with optimization algorithms currently used in either ecosystem management or resource economics, RL algorithms have no guarantees of or metrics for convergence to an optimal solution.
In general, one can only assess the performance of these black box methods relative to alternatives.

<!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
In this article, we compare the performance of a DRL solution to that of a management strategy evaluation across potential 'constant mortality' policies.\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘code’’ subdirectory. The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.} 
We consider the problem of devising harvest strategies in a three species ecosystem (Figure 1B).
We show that, somewhat surprisingly, the DRL policy closely resembles a *bang-bang* solution (see Text Box 1).
Bang-bang solutions are known to be optimal for a variety of single-species models, however, there is no a-priori reason why they should be optimal for higher-dimensional systems.
This way, the DRL agent discovers and exploits, as it were, an effective simplification of the control problem.
We provide some intuition as to how this simplification arises from the dynamics of the system, but leave a more rigorous characterization of this phenomenon for future work.

The DRL policy drives an oscillation process in the system and, in this way, maintains the long-term sustainability of fishing.
In contrast, constant mortality policies overfish at low populations and underfish (relative to the DRL solution) at high populations.
This leads to a tradeoff when leveraging constant mortality strategies: higher constant mortality rates lead to high profitability at the expense of a higher probability of overfishing to extinction, while lower constant mortality rates attain stability at the expense of profit.

@@

In this section we introduce some models describing the population dynamics of fisheries.
In general, the class of models that appear in this context are *first order finite difference equations* (potentially stochastic difference equations).
For $n$ species, these models have the general form
\begin{align}
  \label{eq:general model}
  \begin{split}
    \Delta X_t &= f_X(N_t) - M_X X_t\\
    \Delta Y_t &= f_Y(N_t) - M_Y Y_t\\
    &\dots
  \end{split}
\end{align}
where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $\Delta X_t := X_{t+1} - X_t$, $\Delta Y_t := Y_{t+1} - Y_t$, and where $f_i:\mathbb{R}^{n}\to\mathbb{R}$ are arbitrary functions.
Here, $M_i=M_i(N_t)\in\mathbb{R}_+$ is a state-dependent fish mortality arising from harvesting (sometimes referred to as *fishing effort*).
Thus, the term $M_X X_t$ is the total $X$ harvest at time $t$.
We will focus on systems where a single species, $X$ is harvested, i.e. $M_Y = M_Z = \dots = 0$.

@@

Reinforcement learning (RL) is a way of approaching *sequential decision problems* through machine learning.
All applications of RL can be conceptually separated into two parts: an *agent* and an *environment* which the agent interacts with.
That is, the agent takes actions within the environment.
As a result of an agent's action, the environment will transition to a new state and return a numerical *reward* to the agent. (See Fig. 1 in [@rl-intro] for a conceptual description of reinforcement learning algorithms.)
The rewards encode the agent's goal.
The main task of any RL algorithm is then to maximize the cumulative reward received. 
This objective is achieved by aggregating experience in what is called the *training* period and learning from such experience.
<!-- ML: We could cite our deep RL paper here instead of going into further details. Not sure what we are expecting as the audience. -->.

@@

```{r}
#| label: tuning
#| fig.cap : "Result of the tuning procedure for the constant mortality strategy. 
#| (A) Mean episode reward (averaged over 50 episodes) shown as scatter data, and shaded in gray are two standard deviations around each scatter point. Vertical lines correspond to the optimal constant mortality rate $M^* = $ 0.045 (green) and $0.8 M^*=$ 0.036 (violet). 
#| (B) Several constant mortality policies shown for illustrative purposes. Slopes $0.02$, $0.03$, $0.055$ and $0.065$ shown in black; the optimal linear policy $M^*X$ shown in green, while $0.8M^*X$ shown in violet." 
#| out.width : "6.5in"
msy_rewards <- msy_df |>
  group_by(action, rep) |>
  filter(t == max(t)) |> 
  group_by(action) |>
  summarise(mean_reward = mean(reward),
            sd = sd(reward))

@@

In Fig. \ref{fig:state-space}, we perform a back to back visualization of all the policies evaluated---MSY constant mortality, 80% MSY constant mortality, DRL and optimal constant escapement.
The policies are computed at each state along the paths they traverse in the 50 evaluation rounds, and the values are projected on the $X$-$Z$ plane for visualization.
This figure highlights the similarity between the policy found by the DRL agent and the optimal escapement policy.
